Synthesizing Symmetric Lenses

### 2019-02-18 papers/SymmetricLenses/main.tex
updated with dave's suggestions

{\em Lenses} are programs that can be run both ``front to back'' and ``back
to front,'' allowing updates to both input and output data to be transferred
in both directions.
Since their introduction by~\citet{Focal2005-long}, lenses have been extensively
studied, extended, and applied.  Recent work~\cite{optician} has also demonstrated how
techniques from {\em type-directed program synthesis} can be used to
efficiently synthesize a simple class of lenses---so-called {\em
  bijective lenses} over string data---given a pair of types (regular
expressions) and a small number of examples.

\hspace*{1em}We show how to extend this synthesis algorithm to a much wider class of
lenses, which we call {\em simple symmetric lenses}, including not only all
of the bijective lenses and the more widely used ``asymmetric''
lenses~\cite{Focal2005-long}, but also a rich subset of the full-blown
``symmetric lenses'' proposed by \citet{symmetric-lenses}.

Synthesizing simple symmetric lenses is substantially more challenging than
synthesizing bijective lenses: Since some of the information on each side
can be ``disconnected'' from the other side, there will in general be {\em
  many} lenses that agree with a given example. To guide the search process,
we use {\em stochastic regular expressions} and ideas from information
theory to estimate the amount of information propagated by a candidate lens,
in conjunction with user annotations identifying parts of the source and
target data structures as either {\em irrelevant} or {\em essential}.

@@

We describe an implementation of simple symmetric lenses and our synthesis
procedure as extensions to the Boomerang language~\cite{Boomerang}. We evaluate
its performance on 48 benchmark examples drawn from Flash Fill~\cite{flashfill},
Augeus~\cite{augeas2}, the bidirectional programming literature, and electronic
file format synchronization tasks. Our implementation can synthesize all these
lenses in under 30 seconds.
\end{abstract}

@@

One convenient way to maintain consistency between such varied data formats
is to use a \emph{lens}~\cite{Focal2005-long}---a bidirectional program that
can transform updates to data represented in a format $S$ to another format
$T$ and vice versa, providing ``round tripping'' guarantees that help ensure
information is not lost or corrupted as it is transformed back and forth
between different representations.
Domain-specific languages for writing lenses can facilitate building
principled data transformations.  For example, Boomerang~cite{boomerang} is a
language for expressing bidirectional string transformations.

@@

However, the set of real-world use cases for bijective lenses is also
somewhat limited.  A more common situation is where two real-world data
formats share just {\em some} of their information content.  For instance,
two related database tables might share four out of five columns, but each
might have a different fifth column.  Or, one ad-hoc system configuration
file might include some format-specific meta data, such as a date or a
reference number, whereas the same configuration file on another operating
system does not.  Indeed among the 48 benchmark problems described in
Section \ref{sec:evaluation} below, all of the benchmarks taken from Flash
Fill~\cite{flashfill} and many of the benchmarks that synchronize between
two ad hoc file formats have this characteristic, 13 benchmarks in
total.

@@

The first is pragmatic. ``Classical'' symmetric lenses as presented
by ~\citet{symmetric-lenses} operate
over three structures: the ``left'' structure $X$, the ``right'' $Y$ and a complement
$C$ that contains the information not
present in either $X$ or $Y$. Logistically, these complements must be stored
and managed somehow.  More fundamentially, complements complicate
synthesis {\em specifications}---instead of just giving single examples of
source and target 
pairings, users would have to give longer ``interaction sequences'' to 
show how a lens should behave. To avoid these complexities, we define a
restricted variant 
of symmetric lenses, which we call \emph{simple symmetric lenses}. 
%only the usual source and target.
%
%, which we define in Section~\ref{sec:relationship}.
Intuitively, simple symmetric lenses are the complete set of symmetric
lenses that do not require external ``memory'' to recover data from past
instances of $X$ or $Y$ when making a round trip. They only need the most
recent instance.

@@

We consider a ``likely'' lens to be one that propagates ``a lot'' of information
from the left data format to the right and vice versa. Conversely, an unlikely
lens is one that requires a large amount of additional information to recover
either of the formats given the other. For many application scenarios, users
will want to synchronise formats with lots of shared information (and lenses
with little shared information are typically easier to write). Our tool is built
primarily for these complex scnearios, and so lenses that propagate more
information are preferable in our context to ones that propagate less.
%
This intuition is formalized using \emph{stochastic regular expressions}
(SREs)~\cite{stoch-rnn,stoch-def}, which simultaneously define a set of
strings and a probability distribution over those strings.  Using this
probability distribution, we can calculate the shared information content of
two formats and our measure of likelihood.

@@

With simple symmetric lenses and our SRE-based likelihood measure in hand, we
develop a new algorithm for synthesizing likely lenses. At its core, the
algorithm performs a type-directed search between the descriptions of the source
and target formats, measuring success using the likelihood measure. The
interesting complications arise from the need to deal with regular-expression
equivalences. There are infinitely many regular expressions equivalent to a
given one, and the lens returned by a type-directed search will in general
depend on which of the possible representations are chosen for its source and
target formats. Moreover, if Boomerang's type-equivalence rule is disallowed,
there may be {\em no} well-typed lens connecting the source and target formats
selected by the user: in general, the synthesis algorithm has to search through
equivalent regular expression types for which a suitable lens exists. To tame
this complexity, we divide the synthesis algorithm into two communicating search
procedures (Figure~\ref{fig:high-level-algorithm}), following~\citet{optician}.
The first, \Expand, uses rewriting rules to iteratively propose new pairs of
stochastic regular expressions equivalent to the original pair. The second,
\GreedySynth, takes a pair of SREs as input and uses a greedy, type-directed
algorithm to find a simple symmetric lens between them, returning the lens and
its likelihood score to \Expand. The whole synthesis algorithm heuristically
terminates when the most likely lens is no longer unlikely enough to justify
the time cost of searching through newly generated SREs.

@@

We have added this synthesis procedure to the Boomerang system and evaluated
its effectiveness on a benchmark suite of 48 problems, including data-cleaning
tasks taken from Flash Fill~\cite{gulwani-popl-2014}, view-maintenance tasks
taken from Augeas~\cite{augeas2}, and ad-hoc synchronization tasks from a
variety of other sources (mostly from the bidirectional programming literature).
The user can control the search process via examples and annotations
that informs the synthesis algorithm as to whether certain bits of information
are irrelevant, or must maintained.

In summary, our contributions are:
\begin{itemize}
%
\item We define {\em simple symmetric lenses}, a subclass of symmetric
lenses that do not maintain any hidden state (\S\ref{sec:overview}).  
%
We present a collection of combinators for building simple symmetric lenses and
describe how to calculate their likelihoods (\S\ref{sec:ssl}).
\item We give an algorithm for synthesizing lenses built from these combinators,
  using a novel application of stochastic regular expressions to guide the
  search process (\S\ref{sec:synthesis}) and prove that our metrics have deep
  semantic meaning.
%
\item We extend the Boomerang implementation with simple symmetric lenses,
  demonstrating that they integrate smoothly with other lens extensions such as
  quotient lenses~\cite{quotientlenses} and matching
  lenses~\cite{matchinglenses}. We evaluate our implementation, the effects of
  various optimizations, and our inclusion of relevance annotations on a set of
  48 lens synthesis benchmarks drawn from data cleaning, view maintenance, and
  file synchronization tasks. The system can synthesize simple symmetric lenses
  for all of the benchmarks in under 30 seconds (\S\ref{sec:evaluation}).
%
\item We show that simple symmetric lenses fall between asymmetric
lenses~\cite{Focal2005-long} and full symmetric
lenses~\cite{symmetric-lenses} in expressiveness.  Indeed, the
class of simple lenses can be characterized semantically as the subset of
full symmetric lenses for which a simple ``forgetfulness'' property holds
(\S\ref{sec:related}).
\end{itemize}
Along the
way, we show that {\em star-semiring equivalences} on regular expressions
remain valid when extended to stochastic regular expressions, preserving
probability distributions as well as languages (\S\ref{sec:sre}).
We close with related work (\S\ref{sec:related}) and
concluding thoughts (\S\ref{sec:conc}).
Elided proofs can be found in the appendices
\ifappendices\else included with the supplemental material\fi.

>>>>

{\em Lenses} are programs that can be run both ``front to back'' and ``back
to front,'' allowing updates to both input and output data to be transferred
in both directions.
Since their introduction by Foster \ETAL, lenses have been extensively
studied, extended, and applied.  Recent work has also demonstrated how
techniques from {\em type-directed program synthesis} can be used to
efficiently synthesize a simple class of lenses---so-called {\em
  bijective lenses} over string data---given a pair of types (regular
expressions) and a small number of examples.

\hspace*{1em}We show how to extend this synthesis algorithm to a much wider class of
lenses, which we call {\em simple symmetric lenses}, including not only all
of the bijective lenses and the more widely used ``asymmetric''
lenses, but also a rich subset of the full-blown
``symmetric lenses'' proposed by Hofmann \ETAL.

Synthesizing simple symmetric lenses is substantially more challenging than
synthesizing bijective lenses: Since some of the information on each side
can be ``disconnected'' from the other side, there will, in general, be {\em
  many} lenses that agree with a given example. To guide the search process,
we use {\em stochastic regular expressions} and ideas from information
theory to estimate the amount of information propagated by a candidate lens,
in conjunction with user annotations identifying parts of the source and
target data structures as either {\em irrelevant} or {\em essential}.

@@

We describe an implementation of simple symmetric lenses and our synthesis
procedure as extensions to the Boomerang language. We evaluate
its performance on 48 benchmark examples drawn from Flash Fill,
Augeus, the bidirectional programming literature, and electronic
file format synchronization tasks. Our implementation can synthesize all these
lenses in under 30 seconds.
\end{abstract}

@@

One convenient way to maintain consistency between such varied data formats
is to use a \emph{lens}~\cite{Focal2005-long}---a bidirectional program that
can transform updates to data represented in a format $S$ to another format
$T$ and vice versa, providing ``round tripping'' guarantees that help ensure
information is not lost or corrupted as it is transformed back and forth
between different representations.
Domain-specific languages for writing lenses can facilitate generation of
principled data transformations.  For example, Boomerang~\cite{boomerang} is a
language for expressing bidirectional string transformations.

@@

However, the set of real-world use cases for bijective lenses is also
somewhat limited.  A more common situation is where two real-world data
formats share just {\em some} of their information content.  For instance,
two related database tables might share four out of five columns, but each
might have a different fifth column.  Or, one ad-hoc system configuration
file might include some format-specific metadata, such as a date or a
reference number, whereas the same configuration file on another operating
system does not.  Indeed among the 48 benchmark problems described in
Section \ref{sec:evaluation} below, all of the benchmarks taken from Flash
Fill~\cite{flashfill} and many of the benchmarks that synchronize between
two ad hoc file formats have this characteristic, 13 benchmarks in
total.

@@

The first is pragmatic. ``Classical'' symmetric lenses as presented
by ~\citet{symmetric-lenses} operate
over three structures: the ``left'' structure $X$, the ``right'' $Y$ and a complement
$C$ that contains the information not
present in either $X$ or $Y$. These complements must be stored
and managed somehow.  More fundamentially, complements complicate
synthesis {\em specifications}---instead of just giving single examples of
source and target 
pairings, users would have to give longer ``interaction sequences'' to 
show how a lens should behave. To avoid these complexities, we define a
restricted variant 
of symmetric lenses, which we call \emph{simple symmetric lenses}. 
%only the usual source and target.
%
%, which we define in Section~\ref{sec:relationship}.
Intuitively, simple symmetric lenses are the complete set of symmetric
lenses that do not require external ``memory'' to recover data from past
instances of $X$ or $Y$ when making a round trip. They only need the most
recent instance.

@@

We consider a ``likely'' lens to be one that propagates ``a lot'' of information
from the left data format to the right and vice versa. Conversely, an unlikely
lens is one that requires a large amount of additional information to recover
either of the formats given the other. For many application scenarios, users
will want to synchronise formats with lots of shared information (and lenses
with little shared information are typically easier to write). Our tool is built
primarily for these complex scnearios, and so lenses that propagate more
information are preferable in our context to ones that propagate less.
%
This intuition is formalized using \emph{stochastic regular expressions}
(SREs)~\cite{stoch-rnn,stoch-def}, which simultaneously define a set of
strings and a probability distribution over those strings.  Using this
probability distribution, we can calculate the likelihood of a given lens.

@@

With simple symmetric lenses and our SRE-based likelihood measure in hand, we
develop a new algorithm for synthesizing likely lenses. At its core, the
algorithm performs a type-directed search between the descriptions of the source
and target formats, measuring success using the likelihood measure. The
interesting complications arise from the need to deal with regular-expression
equivalences. There are infinitely many regular expressions equivalent to a
given one, and the lens returned by a type-directed search will in general
depend on which of the possible representations are chosen for its source and
target formats. Moreover, if Boomerang's type-equivalence rule is disallowed,
there may be {\em no} well-typed lens connecting the source and target formats
selected by the user: in general, the synthesis algorithm has to search through
equivalent regular expression types to find the most likely lens. To tame
this complexity, we divide the synthesis algorithm into two communicating search
procedures (Figure~\ref{fig:high-level-algorithm}), following~\citet{optician}.
The first, \Expand, uses rewriting rules to iteratively propose new pairs of
stochastic regular expressions equivalent to the original pair. The second,
\GreedySynth, takes a pair of SREs as input and uses a greedy, type-directed
algorithm to find a simple symmetric lens between them, returning the lens and
its likelihood score to \Expand. The whole synthesis algorithm heuristically
terminates when a sufficiently likely lens is found.

@@

We added this synthesis procedure to the Boomerang system and explored its
effectiveness on a range of applications. Users set up a Boomerang synthesis
task by providing the source and target types and optionally supplying one or
more examples. In addition, users can override the default mechanism for
calculating the information content of a SRE by
asserting that certain strings are \emph{essential} or \emph{irrelevant},
thereby forcing certain data to either be retained or discarded during the
transformations. We evaluated our system on a benchmark suite of 48 problems, including data-cleaning
tasks taken from Flash Fill~\cite{gulwani-popl-2014}, view-maintenance tasks
taken from Augeas~\cite{augeas2}, and ad-hoc synchronization tasks from a
variety of other sources (mostly from the bidirectional programming literature).

In summary, our contributions are:
\begin{itemize}
%
\item We define {\em simple symmetric lenses}, a subclass of symmetric
lenses that do not maintain any hidden state (\S\ref{sec:overview}).  
%
We present a collection of combinators for building simple symmetric lenses and
describe how to calculate their likelihoods (\S\ref{sec:ssl}).
\item We give an algorithm for synthesizing lenses built from these combinators,
  using a novel application of stochastic regular expressions to guide the
  search process (\S\ref{sec:synthesis}) and prove that our metrics have deep
  semantic meaning.
%
\item We extend the Boomerang implementation with simple symmetric lenses,
  demonstrating that they integrate smoothly with other lens extensions such as
  quotient lenses~\cite{quotientlenses} and matching
  lenses~\cite{matchinglenses}.

\item We also extend Boomerang with symmetric lens synthesis specifications,
  which include types, examples, and relevance annotations. We evaluate our implementation, the effects of
  various optimizations, and our inclusion of relevance annotations on a set of
  48 lens synthesis benchmarks drawn from data cleaning, view maintenance, and
  file synchronization tasks. The system can synthesize simple symmetric lenses
  for all of the benchmarks in under 30 seconds (\S\ref{sec:evaluation}).
%
\item We show that simple symmetric lenses fall between asymmetric
lenses~\cite{Focal2005-long} and full symmetric
lenses~\cite{symmetric-lenses} in expressiveness.  Indeed, the
class of simple lenses can be characterized semantically as the subset of
full symmetric lenses for which a simple ``forgetfulness'' property holds
(\S\ref{sec:related}).
\end{itemize}
Along the
way, we show that {\em star-semiring equivalences} on regular expressions
remain valid when extended to stochastic regular expressions, preserving
probability distributions as well as languages (\S\ref{sec:sre}).
We close with related work (\S\ref{sec:related}) and
concluding thoughts (\S\ref{sec:conc}).
Elided proofs can be found in the appendices
\ifappendices\else included with the supplemental material\fi.
